Structural analysis

### 2025-03-22 workflow/notebooks/draft.Rmd
added figs3 in draft

## Fig1

@@

## Fig2

@@

fig2_first_row <- (pident_p | eval_p | qcov_p)
fig2_second_row <- ((lddt_p / alntmscore_p / plddt_p) | p_dbs) + 
  plot_layout(guides = "collect") & 
  theme(legend.position = 'top')
fig2 <- (fig2_first_row / fig2_second_row) +
  plot_layout(heights = c(1,1.3)) +
  plot_annotation(tag_levels = 'A') &
  theme(plot.tag = element_text(size = 10, face = "bold", family = "Aptos"))

@@

## Fig3

@@

# ML trees with selected model and LL
df_ml <- read_delim(paste0(params$outdir, 'phylogeny/',
                              params_dataset$phylo_dataset,  "/trees/", 
                              params_dataset$seed, "_mltrees.txt"),
                    delim = "\t",
                    show_col_types=FALSE) %>%
  # mutate(fn=gsub("_mltrees.txt", "", basename(fn))) %>%
  # separate(fn, c("gene", "target", "alphabet", "model"), sep = "_") %>%
  mutate(model=factor(gsub("3DI", "3Di",
                           gsub("GTR20", "GTR",
                                gsub("\\+.*", "", Model))), levels=models),
         rate=sub("\\+", "", gsub("GTR20|3DI|LG", "", Model)))

@@

```{r FigS2}
# Compare trees with RF distance
rf_df <- get_rf_df(df_trees)
rf_text <- group_by(rf_df, model.x, model.y) %>% 
  # filter(as.integer(model.x)<=as.integer(model.y)) %>% 
  summarise(value=mean(RF, na.rm=T))

plot_bs <- df_bs %>% 
  filter(model!="FM") %>% 
  mutate(targets=target,
         model=factor(model, levels=models)) %>% 
  ggplot(aes(model, support, fill=targets)) +
  geom_boxplot(outlier.size = .1) +
  scale_fill_manual(values = palettes_method) +
  theme_classic() +
  labs(y="Boostrap", x="", fill="") +
  theme(legend.position = "bottom")

@@

rf_plot <- rf_text %>% 
  ggplot(aes(model.x, model.y)) +
  geom_tile(aes(fill=value)) +
  geom_text(aes(label=round(value, 2)), color="white") +
  scale_fill_distiller(palette="YlGnBu", direction = -1, limits = c(0,1), 
                       name="Mean RF") + 
  labs(x="Mean RF Distance", y="") +
  theme_classic() +
  theme(legend.position = "none", 
        axis.line = element_blank(),
        axis.ticks = element_blank())

@@

plot_dist <- df_dist_summ %>% 
  ggplot(aes(median_dist, model,  fill=singleton)) + 
  # ggdist::stat_halfeye(alpha=.8, point_colour = NA, .width = 0, height=.6, 
  #                      position = position_nudge(y=.15)) +
  geom_boxplot(
    # width=.2,
    # position = position_dodge(),
    # removing outliers
    outlier.color = NA,
    alpha = 1,
    # position= position_nudge(y=-.5)
  ) + 
  labs(x="Normalised internode distance", y="Method", fill="") +
  scale_x_continuous(expand = c(0,0.01), limits = c(0,1)) +
  theme(panel.border = element_rect(fill = NA)) + 
  scale_fill_manual(values = palette_singleton_p) + 
  theme(legend.position = "bottom")

fig_s2 <- ((plot_bs / rf_plot) + 
             plot_layout(heights = c(2,1)) | 
             (plot_dist / var_r2t) + 
             plot_layout(heights = c(2,1))) +
  plot_layout(widths = c(1, 2)) +
  plot_annotation(tag_levels = 'A') & 
  theme(plot.tag = element_text(size = 10, face = "bold", family = "Helvetica"))

ggsave("draft/figures/Fig_S2.pdf", fig_s2, width = 10, height = 6)

@@

### Experimental vs predicted

@@

fig_CXX <- (RF_cxx + apro_CXX + DL_cxx + plot_mem_cxx + plot_time_cxx)
fig_CXX <-  fig_CXX +
  plot_layout(guides = "collect", nrow = 1, 
              widths = c(5,5,5,3,3)) +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "none")

>>>>

## Fig 1

@@

## Fig 2

@@

fig2_first_row <- (pident_p | eval_p | qcov_p)
fig2_second_row <- ((lddt_p / alntmscore_p / plddt_p) | p_dbs) + 
  plot_layout(guides = "collect") & 
  theme(legend.position = 'top')
fig2 <- (fig2_first_row / fig2_second_row) +
  plot_layout(heights = c(1,1.3)) +
  plot_annotation(tag_levels = 'A') &
  theme(plot.tag = element_text(size = 10, face = "bold", family = "Helvetica"))

@@

## Fig 3

@@

# ML trees with selected model and LL
mltrees <- paste0(params$outdir, 'phylogeny/',
                              params_dataset$phylo_dataset,  "/trees/", 
                              params_dataset$seed, "_mltrees.txt")
mltrees <- "test/flashforward/out/mltrees.txt"

df_ml <- read_delim(mltrees,
                    delim = "\t",
                    show_col_types=FALSE) %>%
  # mutate(fn=gsub("_mltrees.txt", "", basename(fn))) %>%
  # separate(fn, c("gene", "target", "alphabet", "model"), sep = "_") %>%
  mutate(model=factor(gsub("3DI", "3Di",
                           gsub("GTR20", "GTR",
                                gsub("\\+.*", "", Model))), levels=models),
         rate=sub("\\+", "", gsub("GTR20|3DI|LG", "", Model)))

@@

## Fig S1

can be found here: "results/ortho_bench/eggnog/accuracy.pdf" and can be reproduced
with any phylogenomics database annotated in uniprot following the instructions
in the main README.md

## Fig S2

```{r FigS2}
# Compare trees with RF distance
rf_df <- get_rf_df(filter(df_trees, model %in% models_draft))
rf_text <- group_by(rf_df, model.x, model.y) %>% 
  # filter(as.integer(model.x)<=as.integer(model.y)) %>% 
  summarise(value=mean(RF, na.rm=T))

plot_bs <- df_bs %>% 
  filter(model %in% c("LG", "GTR", "3Di", "AF", "LLM")) %>% 
  mutate(targets=target,
         model=factor(model, levels=models)) %>% 
  ggplot(aes(model, support, fill=targets)) +
  geom_boxplot(outlier.size = .1) +
  scale_fill_manual(values = palettes_method) +
  theme_classic() +
  labs(y="Boostrap", x="", fill="") +
  theme(legend.position = "bottom")

@@

rf_plot <- rf_text %>% 
  ggplot(aes(model.x, model.y)) +
  geom_tile(aes(fill=value)) +
  geom_text(aes(label=round(value, 2)), color="white", size=3) +
  scale_fill_distiller(palette="YlGnBu", direction = -1, limits = c(0,1), 
                       name="Mean RF") + 
  labs(x="Mean RF Distance", y="") +
  theme_classic() +
  theme(legend.position = "none", 
        axis.line = element_blank(),
        axis.ticks = element_blank())

plot_struct_ml <- df_ml %>% 
  filter(model %in% c("GTR", "3Di", "AF", "LLM")) %>% 
  select(gene, targets, model, AIC, AICc, BIC) %>% 
  pivot_longer(!c(gene, targets, model)) %>% 
  group_by(gene, targets, name) %>% 
  mutate(rank = dense_rank(value)) %>% 
  filter(rank==1) %>% 
  ggplot(aes(model)) + 
  geom_bar(fill="grey40") + 
  scale_y_continuous(expand = expansion(mult = c(0,.02))) +
  facet_grid(~name) + 
  labs(x="", y="N. first in metric")

@@

plot_dist <- df_dist_summ %>% 
    mutate(model=factor(model, levels=models)) %>% 
  ggplot(aes(model, median_dist,  fill=singleton)) + 
  geom_boxplot(
    outlier.color = NA,
    alpha = 1
  ) + 
  labs(y="Normalised internode distance", y="Method", fill="") +
  scale_y_continuous(expand = c(0,0.01), limits = c(0,1)) +
  theme(panel.border = element_rect(fill = NA)) + 
  scale_fill_manual(values = palette_singleton_p) + 
  theme(legend.position = "bottom")

first_row <- plot_bs + rf_plot + plot_struct_ml 
second_row <- plot_dist + var_r2t

fig_s2 <- (first_row / second_row) +
  plot_annotation(tag_levels = 'A') &
  theme(plot.tag = element_text(size = 10, face = "bold", family = "Helvetica"))

# fig_s2 <- ((plot_bs / rf_plot) + 
#              plot_layout(heights = c(2,1)) | 
#              (plot_dist / var_r2t) + 
#              plot_layout(heights = c(2,1))) +
#   plot_layout(widths = c(1, 2)) +
#   plot_annotation(tag_levels = 'A') & 
#   theme(plot.tag = element_text(size = 10, face = "bold", family = "Helvetica"))

ggsave("draft/figures/Fig_S2.pdf", fig_s2, width = 12, height = 8, device = cairo_pdf)

@@

## Fig S3

```{r, figS3}
bench_dir <- "draft/foldtree_benchmark/zenodo/OMA_data/Eukaryota/reconciliation/"
sptree_ft <- read.tree(paste0(bench_dir, "sptree.nwk"))

ts_files_ft <- list.files(paste0(bench_dir, "trees/tmp"), pattern = "*nwk", full.names = T)
ts_ft <- read.tree(text = sapply(ts_files_ft, readLines))
names(ts_ft) <- gsub(".nwk", "", basename(ts_files_ft))

n_tips_ft <- sapply(ts_ft, function(x) length(x$tip.label)) %>% 
  enframe() %>%
  separate(name, c("gene", "mat"), sep = "_") %>%
  rename("n_tips"="value")

TCS_ft <- read_delim(paste0(bench_dir,"TCS.txt"),
           delim = "\t", col_names = c("gene", "mat", "TCS")) %>% 
  mutate(mat = str_trim(mat))

apro_files_ft <- list.files(bench_dir, pattern = "^apro", full.names = T)
apro_trees_ft <- read.tree(text = sapply(apro_files_ft, readLines))
names(apro_trees_ft) <- gsub(".txt", "", sapply(str_split(basename(apro_files_ft), "_"), 
                            function(x) paste0(x[2])))
trees_df_ft <- get_apro_stats(apro_trees_ft)

disco_files_ft <- list.files(bench_dir, pattern = "^disco", full.names = T)
disco_rf_ft <- get_disco_rf(disco_files_ft, sptree_ft) %>% 
  mutate(mat = gsub(".txt", "", targets)) %>% 
  select(-targets, -alphabet, -model)

notung_files_ft <- list.files(bench_dir, pattern = "^notung", full.names = T)[2:5]

notung_ft <- read_delim(notung_files_ft, 
                   show_col_types = FALSE, 
                   col_names = c("gene", "dups", "losses"), 
           id = "mat") %>% 
  mutate(mat = gsub("notung_|.txt", "", basename(mat))) %>% 
  left_join(n_tips_ft) %>% 
  mutate(n_events = (dups+losses)/n_tips)

apro_plot_ft <- trees_df_ft %>%
  mutate(targets = gsub("iq", "iqtree", targets)) %>% 
  ggplot(aes(fct_rev(fct_reorder(targets, q1)), q1)) + 
  geom_boxplot(outlier.size = .1) + 
  labs(x="", y="Quartet support")

disco_plot_ft <- disco_rf_ft %>%
  mutate(mat = gsub("iq", "iqtree", mat)) %>%
  ggplot(aes(fct_reorder(mat, rf), rf)) + 
  geom_boxplot(outlier.size = .1) + 
  labs(y="Norm. RF to species tree", x="")

notung_plot_ft <- notung_ft %>%
  mutate(mat = gsub("iq", "iqtree", mat)) %>% 
  ggplot(aes(fct_reorder(mat, n_events), n_events)) + 
  geom_boxplot(outlier.size = .1) + 
  labs(y="(D+L)/# Tips", x="")

TCS_plot <- TCS_ft %>% 
  ggplot(aes(fct_rev(fct_reorder(mat, TCS)), TCS)) + 
  geom_boxplot(outlier.size = .1) + 
  scale_y_log10() + 
  labs(x="")

TCS_scatter <- TCS_ft %>% 
  filter(mat %in% c("iqtree", "fident")) %>% 
  pivot_wider(names_from = mat, values_from = TCS) %>% 
  mutate(is_better = iqtree > fident) %>% 
  add_count(is_better) %>% 
  mutate(`IQ-Tree Better` = paste0(is_better, ' (', n, ')')) %>% 
  # count(is_better)
  ggplot(aes(iqtree, fident, color=`IQ-Tree Better`)) + 
  geom_point() + 
  geom_abline(slope = 1) + 
  coord_fixed() + 
  labs(x="TCS - IQ-Tree", y="TCS - Fident") +
  scale_color_manual(values = c("blue", "red"))

left_side <- ((apro_plot_ft | disco_plot_ft) / (notung_plot_ft | TCS_plot))
fig_S3 <- (left_side | TCS_scatter) +
  plot_layout(widths = c(1,1)) +
  plot_annotation(tag_levels = 'A') &
  theme(plot.tag = element_text(size = 10, face = "bold", family = "Helvetica"))

ggsave("draft/figures/Fig_S3.pdf", fig_S3, width = 11, height = 5, device = cairo_pdf)
```

## Fig S4

@@

fig_CXX <- (RF_cxx + apro_CXX + DL_cxx + plot_mem_cxx + plot_time_cxx)
fig_CXX <-  fig_CXX +
  plot_layout(guides = "collect", nrow = 1, 
              widths = c(5,5,5,3,3)) +
  plot_annotation(tag_levels = 'A') &
  theme(plot.tag = element_text(size = 10, face = "bold", family = "Helvetica"))
